---
title: "Frequency Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, warning = FALSE, message = FALSE,
    dpi = 144, out.width = "100%", fig.align = "center"
)
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
source("theme_ej.R")

# Read the data (assuming this was already done in data-overview.Rmd)
# If you're using this page independently, uncomment and run the data loading code
results <- read.csv("eurojackpot_results.csv", stringsAsFactors = FALSE)

# Convert date format to day, month, year
results$draw_date <- as.Date(results$draw_date)
results$day <- format(results$draw_date, "%d")
results$month <- format(results$draw_date, "%m")
results$year <- format(results$draw_date, "%Y")

# Split main numbers into separate columns
main_numbers_split <- strsplit(gsub("\"", "", results$main_numbers), ",")
main_numbers_matrix <- do.call(rbind, main_numbers_split)
colnames(main_numbers_matrix) <- paste0("main_", 1:5)
results <- cbind(results, main_numbers_matrix)

# Split euro numbers into separate columns
euro_numbers_split <- strsplit(gsub("\"", "", results$euro_numbers), ",")
euro_numbers_matrix <- do.call(rbind, euro_numbers_split)
colnames(euro_numbers_matrix) <- paste0("euro_", 1:2)
results <- cbind(results, euro_numbers_matrix)

# Convert number columns to numeric
for (col in c(paste0("main_", 1:5), paste0("euro_", 1:2))) {
    results[[col]] <- as.numeric(results[[col]])
}

# Euro pool size per draw (the game changed twice - see Chapter 5)
results$pool_size <- ifelse(results$draw_date < as.Date("2014-10-10"), 8,
    ifelse(results$draw_date < as.Date("2022-03-29"), 10, 12)
)

mains <- as.matrix(results[, paste0("main_", 1:5)])
n <- nrow(results)
```

Every lottery fan has a lucky number, and every lucky number has a track record. This chapter counts how often each ball actually came out of the machine, year by year — then asks the only question that matters: is any of the unevenness bigger than pure chance would produce? We standardise every count into a z-score (how many standard deviations a number sits from its fair-machine expectation) and light up the cells that cross the conventional ±1.96 "significance" line.

## Year-by-Year Deviations From a Fair Machine

Instead of fourteen stacked bar charts, one picture: each cell is a number's deviation from expectation in one year. Blue means "came up less than expected", red means "more", and outlined cells cross the 95% significance line.

```{r main-z, fig.height=6.5}
main_z <- results %>%
    select(year, all_of(paste0("main_", 1:5))) %>%
    pivot_longer(-year, values_to = "number") %>%
    count(year, number, name = "frequency") %>%
    complete(year, number = 1:50, fill = list(frequency = 0)) %>%
    group_by(year) %>%
    mutate(
        draws = sum(frequency) / 5,
        expected = draws * 5 / 50,
        z_score = (frequency - expected) / sqrt(draws * 0.1 * 0.9),
        is_outlier = abs(z_score) > 1.96
    ) %>%
    ungroup()

n_out_main <- sum(main_z$is_outlier)

ggplot(main_z, aes(x = number, y = year, fill = z_score)) +
    geom_tile(color = ej_surface, linewidth = 0.4) +
    geom_tile(
        data = filter(main_z, is_outlier),
        color = ej_ink, linewidth = 0.55
    ) +
    scale_fill_gradient2(
        low = ej_blue_dark, mid = "#f0efec", high = ej_red,
        name = "z-score"
    ) +
    scale_x_continuous(breaks = c(1, seq(5, 50, by = 5)), expand = c(0, 0)) +
    labs(
        title = "Main Numbers: Deviation From Expectation, Year by Year",
        subtitle = sprintf(
            "Outlined cells pass |z| > 1.96. Found: %d of %d cells (%.0f expected as false positives at the 95%% line)",
            n_out_main, nrow(main_z), 0.05 * nrow(main_z)
        ),
        x = "Main number (1-50)", y = NULL, caption = ej_caption
    ) +
    theme(panel.grid.major = element_blank(), legend.key.width = unit(28, "pt"))
```

```{r euro-z, fig.height=5.5}
# For euro numbers the expectation must respect the pool changes:
# 9-10 exist only since Oct 2014, 11-12 since Mar 2022. Cells where a number
# was never in the pool are blanked instead of being scored as "cold".
euro_cells <- expand.grid(
    year = sort(unique(results$year)), number = 1:12,
    stringsAsFactors = FALSE
)
euro_stats <- mapply(function(y, x) {
    sel <- results$year == y & results$pool_size >= x
    p <- 2 / results$pool_size[sel]
    c(expected = sum(p), variance = sum(p * (1 - p)), exposure = sum(sel))
}, euro_cells$year, euro_cells$number)
euro_cells <- cbind(euro_cells, t(euro_stats))

euro_obs_year <- results %>%
    select(year, euro_1, euro_2) %>%
    pivot_longer(-year, values_to = "number") %>%
    count(year, number, name = "frequency")

euro_z <- euro_cells %>%
    left_join(euro_obs_year, by = c("year", "number")) %>%
    mutate(
        frequency = coalesce(frequency, 0L),
        z_score = ifelse(exposure > 0, (frequency - expected) / sqrt(variance), NA),
        is_outlier = !is.na(z_score) & abs(z_score) > 1.96
    )

n_out_euro <- sum(euro_z$is_outlier)
n_cells_euro <- sum(euro_z$exposure > 0)

ggplot(euro_z, aes(x = number, y = year, fill = z_score)) +
    geom_tile(color = ej_surface, linewidth = 0.4) +
    geom_tile(
        data = filter(euro_z, is_outlier),
        color = ej_ink, linewidth = 0.55
    ) +
    scale_fill_gradient2(
        low = ej_blue_dark, mid = "#f0efec", high = ej_red,
        name = "z-score", na.value = "#e8e7e2"
    ) +
    scale_x_continuous(breaks = 1:12, expand = c(0, 0)) +
    labs(
        title = "Euro Numbers: Deviation From Expectation, Year by Year",
        subtitle = sprintf(
            "Gray cells = number not yet in the pool. Outliers: %d of %d live cells (%.0f expected by chance)",
            n_out_euro, n_cells_euro, 0.05 * n_cells_euro
        ),
        x = "Euro number (1-12)", y = NULL, caption = ej_caption
    ) +
    theme(panel.grid.major = element_blank(), legend.key.width = unit(28, "pt"))
```

```{r outlier-table}
main_outliers_summary <- main_z %>%
    filter(is_outlier) %>%
    arrange(desc(abs(z_score))) %>%
    head(10) %>%
    transmute(
        Year = year, `Main number` = number, Appearances = frequency,
        Expected = round(expected, 1), `z-score` = round(z_score, 2)
    )

knitr::kable(main_outliers_summary,
    caption = "The ten loudest single-year deviations - every one of them ordinary sampling noise",
    align = c("c", "c", "r", "r", "r")
)
```

<div class="pull-stat"><span class="big">`r n_out_main + n_out_euro` outliers, ~`r round(0.05 * (nrow(main_z) + n_cells_euro))` expected</span><span class="why">At a 95% significance line, one cell in twenty crosses it by luck alone. The machine is behaving exactly like a fair machine bragging about being fair.</span></div>

## Most and Least Frequent Numbers (Current Rules)

The game's format changed in March 2022, so for a "who's hot right now" table the honest window is the current-rules era only. Here is every number's record since then:

```{r current-freq, fig.height=5}
current <- results %>% filter(draw_date >= as.Date("2022-03-29"))
n_cur <- nrow(current)

main_cur <- tabulate(as.integer(as.matrix(current[, paste0("main_", 1:5)])), nbins = 50)
expected_cur <- 5 * n_cur / 50

ggplot(data.frame(number = 1:50, frequency = main_cur), aes(number, frequency)) +
    geom_col(fill = ej_blue, width = 0.75) +
    geom_hline(yintercept = expected_cur, linetype = "dashed", color = ej_ink, linewidth = 0.6) +
    annotate("text",
        x = 50.6, y = expected_cur, label = sprintf("expected: %.0f", expected_cur),
        hjust = 1, vjust = -0.7, size = 3.4, color = ej_ink2, fontface = "bold"
    ) +
    scale_x_continuous(breaks = c(1, seq(5, 50, by = 5))) +
    labs(
        title = "Main Number Frequency Under the Current Rules",
        subtitle = sprintf("%d draws since the March 2022 relaunch - dashed line is the fair-machine expectation", n_cur),
        x = "Main number (1-50)", y = "Appearances", caption = ej_caption
    )
```

```{r current-euro, fig.height=4.5}
euro_cur <- tabulate(c(current$euro_1, current$euro_2), nbins = 12)
expected_euro_cur <- 2 * n_cur / 12

ggplot(data.frame(number = 1:12, frequency = euro_cur), aes(number, frequency)) +
    geom_col(fill = ej_gold, width = 0.6) +
    geom_hline(yintercept = expected_euro_cur, linetype = "dashed", color = ej_ink, linewidth = 0.6) +
    annotate("text",
        x = 12.4, y = expected_euro_cur, label = sprintf("expected: %.0f", expected_euro_cur),
        hjust = 1, vjust = -0.7, size = 3.4, color = ej_ink2, fontface = "bold"
    ) +
    scale_x_continuous(breaks = 1:12) +
    labs(
        title = "Euro Number Frequency Under the Current Rules",
        subtitle = "All twelve numbers have been in the pool for this whole window - finally a fair comparison",
        x = "Euro number (1-12)", y = "Appearances", caption = ej_caption
    )
```

```{r current-tables}
top5 <- order(-main_cur)[1:5]
bottom5 <- order(main_cur)[1:5]
knitr::kable(
    data.frame(
        Rank = 1:5,
        `Hottest main` = top5, Appearances = main_cur[top5],
        `Coldest main` = bottom5, `Appearances ` = main_cur[bottom5],
        check.names = FALSE
    ),
    caption = sprintf("Main-number extremes since March 2022 (expected: %.0f each)", expected_cur),
    align = rep("c", 5)
)

knitr::kable(
    data.frame(
        `Euro number` = order(-euro_cur),
        Appearances = euro_cur[order(-euro_cur)],
        `vs expected` = sprintf("%+d", euro_cur[order(-euro_cur)] - round(expected_euro_cur)),
        check.names = FALSE
    ),
    caption = sprintf("All twelve euro numbers since March 2022 (expected: %.0f each)", expected_euro_cur),
    align = c("c", "r", "r")
)
```

<div class="info-box">
<strong>What to take away:</strong> the gap between the hottest and coldest number is precisely chance-sized — the formal chi-squared test in <a href="predictive-analysis.html">Chapter 5</a> gives p&nbsp;=&nbsp;0.98, and the outlined heatmap cells above appear at almost exactly the false-positive rate the 95% line predicts. Frequency tables are fun history. They are not a forecast.
</div>

<div class="navigation-buttons">
  <a href="data-overview.html" class="prev-button"><i class="fa fa-arrow-left"></i> Previous: Data Overview</a>
  <a href="time-series.html" class="next-button">Next: Time Series <i class="fa fa-arrow-right"></i></a>
</div>
