---
pagetitle: "Eurojackpot Analysis — The Numbers Behind Europe's €120M Obsession"
output:
  html_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
results <- read.csv("eurojackpot_results.csv", stringsAsFactors = FALSE)
results$draw_date <- as.Date(results$draw_date)
n_draws <- nrow(results)
n_years <- length(unique(format(results$draw_date, "%Y")))
latest <- results[which.max(results$draw_date), ]
latest_main <- sort(as.numeric(strsplit(gsub('"', "", latest$main_numbers), ",")[[1]]))
latest_euro <- sort(as.numeric(strsplit(gsub('"', "", latest$euro_numbers), ",")[[1]]))
odds <- choose(50, 5) * choose(12, 2)
balls_drawn <- n_draws * 7
```

<div class="ej-hero">
<p class="ej-kicker">Fourteen years · `r format(n_draws, big.mark = ",")` draws · one honest question</p>
<h1 class="ej-headline">The numbers behind Europe's <span class="ej-accent">€120&nbsp;million</span> obsession</h1>
<p class="ej-deck">Eurojackpot has been drawing balls in Helsinki twice a week since 2012. This site puts every single draw under a statistical microscope — the patterns, the myths, and the one thing you can actually predict.</p>

```{r balls, results='asis'}
cat('<div class="ej-balls" aria-label="Most recent draw in the dataset">\n')
cat(paste0('<span class="ball">', latest_main, "</span>"), sep = "\n")
cat('<span class="ball plus">+</span>\n')
cat(paste0('<span class="ball euro">', latest_euro, "</span>"), sep = "\n")
cat("</div>\n")
```

<p class="ej-balls-caption">The most recent draw in this dataset — `r format(latest$draw_date, "%B %e, %Y")`</p>

<div class="stat-grid">
<div class="stat-card"><div class="stat-value"><span class="count-up" data-target="`r n_draws`">`r format(n_draws, big.mark = ",")`</span></div><div class="stat-label">Draws analysed</div></div>
<div class="stat-card"><div class="stat-value"><span class="count-up" data-target="`r balls_drawn`">`r format(balls_drawn, big.mark = ",")`</span></div><div class="stat-label">Balls drawn</div></div>
<div class="stat-card"><div class="stat-value"><span class="count-up" data-target="`r n_years`">`r n_years`</span></div><div class="stat-label">Years covered</div></div>
<div class="stat-card"><div class="stat-value">1<span class="stat-unit"> in </span><span class="count-up" data-target="`r odds`">`r format(odds, big.mark = ",")`</span></div><div class="stat-label">Jackpot odds</div></div>
</div>
</div>

## The game, in one paragraph

Pick **5 numbers from 50**, add **2 Euro numbers from 12**, and match all seven to win a jackpot that starts at €10 million and climbs to a €120 million cap. Draws happen every Tuesday and Friday in Helsinki, across 18 countries. Since launch the game has quietly changed its own rules twice — a detail that, as you'll see, quietly breaks most casual analyses of it.

<div class="pull-stat"><span class="big">1 in 139,838,160</span><span class="why">Your odds of hitting the jackpot with a single line. You are roughly 100× more likely to be struck by lightning this year.</span></div>

## Inside the data

Five chapters, each built from the raw draw history with reproducible R code — press any "Code" button on the way to see exactly how a figure was made.

<div class="page-cards">

<a class="page-card" href="data-overview.html"><span class="pc-kicker">Chapter 1</span><span class="pc-title">Data Overview</span><span class="pc-desc">The shape of `r format(n_draws, big.mark = ",")` draws: what the dataset holds, how often each number shows up, and which pairs travel together.</span><span class="pc-go">Read chapter →</span></a>

<a class="page-card" href="frequency-analysis.html"><span class="pc-kicker">Chapter 2</span><span class="pc-title">Frequency Analysis</span><span class="pc-desc">"Hot" and "cold" numbers under a microscope — and how big the gap between luckiest and unluckiest ball should be by pure chance.</span><span class="pc-go">Read chapter →</span></a>

<a class="page-card" href="time-series.html"><span class="pc-kicker">Chapter 3</span><span class="pc-title">Time Series</span><span class="pc-desc">Draw sums, spreads and seasons: does the machine drift over fourteen years, or is every wobble just noise wearing a trend costume?</span><span class="pc-go">Read chapter →</span></a>

<a class="page-card" href="number-patterns.html"><span class="pc-kicker">Chapter 4</span><span class="pc-title">Number Patterns</span><span class="pc-desc">The gaps between the five balls in a draw — how tightly they cluster, how often neighbours appear, and what a "typical" ticket looks like.</span><span class="pc-go">Read chapter →</span></a>

<a class="page-card" href="predictive-analysis.html"><span class="pc-kicker">Chapter 5</span><span class="pc-title">Predicting the Next Pot?</span><span class="pc-desc">Every popular prediction strategy, formally backtested against history — plus the one quantity in this game you genuinely <em>can</em> forecast.</span><span class="pc-go">Read chapter →</span></a>

</div>

## How we work

Descriptive statistics, formal hypothesis tests, and walk-forward backtests — visualised, and honest about uncertainty. Where a popular belief fails a test, we say so; where the data genuinely contains structure (it happens!), we show where it comes from.

<div class="info-box">
<strong>Play-it-straight note:</strong> lottery draws are independent random events. Past results cannot predict future outcomes, and this site will show you precisely <em>why</em> — with tests, not slogans. Everything here is for education and curiosity. If you play, treat the ticket as entertainment, never as an investment.
</div>

<div class="navigation-buttons">
  <a href="data-overview.html" class="next-button">Start reading: Data Overview <i class="fa fa-arrow-right"></i></a>
</div>
