---
title: "Number Patterns Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, warning = FALSE, message = FALSE,
    dpi = 144, out.width = "100%", fig.align = "center"
)
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
source("theme_ej.R")

# Read the data (assuming this was already done in data-overview.Rmd)
# If you're using this page independently, uncomment and run the data loading code
results <- read.csv("eurojackpot_results.csv", stringsAsFactors = FALSE)

# Convert date format to day, month, year
results$draw_date <- as.Date(results$draw_date)
results$day <- format(results$draw_date, "%d")
results$month <- format(results$draw_date, "%m")
results$year <- format(results$draw_date, "%Y")

# Split main numbers into separate columns
main_numbers_split <- strsplit(gsub("\"", "", results$main_numbers), ",")
main_numbers_matrix <- do.call(rbind, main_numbers_split)
colnames(main_numbers_matrix) <- paste0("main_", 1:5)
results <- cbind(results, main_numbers_matrix)

# Split euro numbers into separate columns
euro_numbers_split <- strsplit(gsub("\"", "", results$euro_numbers), ",")
euro_numbers_matrix <- do.call(rbind, euro_numbers_split)
colnames(euro_numbers_matrix) <- paste0("euro_", 1:2)
results <- cbind(results, euro_numbers_matrix)

# Convert number columns to numeric
for (col in c(paste0("main_", 1:5), paste0("euro_", 1:2))) {
    results[[col]] <- as.numeric(results[[col]])
}

n <- nrow(results)
```

What does a *typical* winning ticket look like? Not the numbers themselves — the shape of them: how far apart the five balls land, how often two neighbours show up together, whether the machine prefers tight clusters or wide scatters. This chapter measures the geometry of `r format(n, big.mark = ",")` draws — and, as usual, checks every "pattern" against what randomness predicts before anyone gets excited.

## Gaps Between Consecutive Numbers

Sort a draw's five main numbers and measure the four gaps between neighbours. The outer gaps (1→2, 4→5) can stretch wide, while the middle gaps stay tighter — a classic signature of order statistics, not of the machine's mood:

```{r consecutive-diffs, fig.height=5.5}
# Calculate pairwise differences between consecutive numbers
main_diffs <- results %>%
    mutate(
        diff_1_2 = main_2 - main_1,
        diff_2_3 = main_3 - main_2,
        diff_3_4 = main_4 - main_3,
        diff_4_5 = main_5 - main_4
    ) %>%
    select(draw_date, year, contains("diff_"))

main_diffs_long <- main_diffs %>%
    pivot_longer(
        cols = starts_with("diff_"),
        names_to = "pair",
        values_to = "difference"
    ) %>%
    mutate(pair = factor(pair,
        levels = c("diff_1_2", "diff_2_3", "diff_3_4", "diff_4_5"),
        labels = c("gap 1 → 2", "gap 2 → 3", "gap 3 → 4", "gap 4 → 5")
    ))

ggplot(main_diffs_long, aes(x = difference)) +
    geom_histogram(binwidth = 1, fill = ej_blue, color = ej_surface, linewidth = 0.2) +
    facet_wrap(~pair, ncol = 2) +
    labs(
        title = "How Far Apart Are Neighbouring Numbers in a Draw?",
        subtitle = "All draws since 2012, split by which gap of the sorted ticket is measured",
        x = "Gap between the two numbers", y = "Number of draws", caption = ej_caption
    )
```

<div class="pull-stat"><span class="big">`r sprintf("%.1f%%", 100 * mean(apply(as.matrix(results[, paste0("main_", 1:5)]), 1, function(m) any(diff(sort(m)) == 1))))` of draws</span><span class="why">contain at least two directly consecutive numbers (like 27 and 28). Theory predicts `r sprintf("%.1f%%", 100 * (1 - choose(46, 5) / choose(50, 5)))` — so stop being surprised when it happens.</span></div>

## Euro Number Gaps

The same idea for the two euro numbers, with the theoretical distribution overlaid — for a pool of size $p$, a gap of $d$ can be formed $p-d$ ways, and the pool size changed twice:

```{r euro-diffs, fig.height=4.5}
results$pool_size <- ifelse(results$draw_date < as.Date("2014-10-10"), 8,
    ifelse(results$draw_date < as.Date("2022-03-29"), 10, 12)
)
euro_gap <- data.frame(gap = results$euro_2 - results$euro_1)

expected_gap <- sapply(1:11, function(d) {
    sum(ifelse(results$pool_size > d,
        (results$pool_size - d) / choose(results$pool_size, 2), 0
    ))
})

ggplot(euro_gap, aes(x = gap)) +
    geom_histogram(aes(fill = "Observed"), binwidth = 1, color = ej_surface, linewidth = 0.3) +
    geom_point(
        data = data.frame(gap = 1:11, expected = expected_gap),
        aes(y = expected, color = "Expected (pool-size adjusted)"), size = 3
    ) +
    scale_fill_manual(values = c("Observed" = ej_gold)) +
    scale_color_manual(values = c("Expected (pool-size adjusted)" = ej_ink2)) +
    scale_x_continuous(breaks = 1:11) +
    labs(
        title = "Gap Between the Two Euro Numbers",
        subtitle = "Small gaps are simply more numerous - the machine agrees with the counting argument",
        x = "Gap between euro numbers", y = "Number of draws", caption = ej_caption
    )
```

## Yearly Trends in Number Gaps

If the machine's geometry drifted, the average gaps would trend. They don't — they braid around their expectations:

```{r yearly-trends, fig.height=5}
yearly_avg_diffs <- main_diffs_long %>%
    group_by(year, pair) %>%
    summarise(
        avg_diff = mean(difference, na.rm = TRUE),
        .groups = "drop"
    )

ggplot(yearly_avg_diffs, aes(x = year, y = avg_diff, color = pair, group = pair)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2.1) +
    scale_color_manual(values = ej_ramp(4)) +
    labs(
        title = "Average Gap Between Neighbouring Numbers, by Year",
        subtitle = "Four ordered gaps, fourteen years, zero direction",
        x = NULL, y = "Average gap", caption = ej_caption
    ) +
    guides(color = guide_legend(nrow = 1)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## The Gap Fingerprint

All four gaps share one fingerprint: short distances dominate and long ones thin out geometrically. The heatmap shows each gap position's distribution as a column of probabilities:

```{r heatmap-alltime, fig.height=5.5}
heatmap_data <- main_diffs_long %>%
    group_by(pair) %>%
    count(difference) %>%
    mutate(percentage = n / sum(n) * 100) %>%
    ungroup()

ggplot(heatmap_data, aes(x = pair, y = difference, fill = percentage)) +
    geom_tile(color = ej_surface, linewidth = 0.3) +
    scale_fill_gradient(
        low = ej_blue_pale, high = ej_blue_deep,
        name = "Share of draws (%)"
    ) +
    scale_y_continuous(breaks = seq(0, 45, by = 5), expand = c(0, 0)) +
    labs(
        title = "Distribution of Each Gap of the Sorted Ticket",
        subtitle = "Darker = more common. Every gap loves small values; the tail thins out fast",
        x = NULL, y = "Gap size", caption = ej_caption
    ) +
    theme(panel.grid.major = element_blank(), legend.key.width = unit(28, "pt"))
```

```{r heatmap-yearly, fig.height=8}
heatmap_yearly <- main_diffs_long %>%
    group_by(pair, year) %>%
    count(difference) %>%
    mutate(percentage = n / sum(n) * 100) %>%
    ungroup()

ggplot(heatmap_yearly, aes(x = pair, y = difference, fill = percentage)) +
    geom_tile() +
    scale_fill_gradient(
        low = ej_blue_pale, high = ej_blue_deep,
        name = "Share of draws (%)"
    ) +
    facet_wrap(~year) +
    labs(
        title = "The Same Fingerprint, Every Single Year",
        subtitle = "Fourteen re-runs of the experiment produce fourteen near-identical patterns",
        x = NULL, y = "Gap size", caption = ej_caption
    ) +
    theme(
        panel.grid.major = element_blank(),
        legend.key.width = unit(28, "pt"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4, size = rel(0.7))
    )
```

<div class="info-box">
<strong>Reading patterns responsibly:</strong> everything above — the gap shapes, the consecutive-number rate, the yearly fingerprints — matches the combinatorics of drawing 5 balls from 50 without replacement. The patterns are real, but they belong to <em>mathematics</em>, not to the machine's habits. A ticket shaped like a "typical" draw wins exactly as often as any other ticket.
</div>

<div class="navigation-buttons">
  <a href="time-series.html" class="prev-button"><i class="fa fa-arrow-left"></i> Previous: Time Series</a>
  <a href="predictive-analysis.html" class="next-button">Next: Predicting the Next Pot? <i class="fa fa-arrow-right"></i></a>
</div>
